---
title: "Project Proposal"
author: "Alec Neme (84761030)"
date: "2026-03-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyquant) 
library(dplyr)
library(lubridate)
```

Which assumption—discrete regime shifts (HMM) or continuous volatility evolution (Stochastic Volatility)-better explains the market dynamics during the Dot-com bubble, the 2008 Financial Crisis and the COVID-19 crisis?

- ^GSPC (S&P 500 Index): Top 500 stocks in the U.S. (probably should use this).

- ^VIX (CBOE Volatility Index): Market volatility measure that we can use for model validation.

- XLK (Technology Select Sector SPDR Fund): An ETF containing all the tech stocks in the S&P 500. Maybe use for dot com crash? 

- XLF (Financial Select Sector SPDR Fund): An ETF containing all the financial stocks in the S&P 500 (banks, insurance, etc.). Check for 2008 Financial Crisis?

- TLT (iShares 20+ Year Treasury Bond ETF): This tracks long-term US government bonds. "Safe haven" asset that normally behaves opposite to stocks. Does the model still work with this IDK?

We use log return $log(P_t / P_{t-1})$ because it makes it easier to work with financial data.

```{r}
project_tickers <- c("^GSPC", "^VIX", "XLK", "XLF", "TLT")
market_data <- tq_get(project_tickers,
                      get  = "stock.prices",
                      from = "1998-01-01")

weekly_returns <- market_data %>%
  mutate(week = floor_date(date, "week")) %>%
  group_by(symbol, week) %>%
  arrange(date) %>%
  slice_tail(n = 1) %>%
  ungroup()
  
  
weekyly_log_returns <- weekly_returns %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(
    previous_price = lag(adjusted),
    log_return = log(adjusted / previous_price)
  ) %>%
  
  filter(!is.na(log_return)) %>%
  select(symbol, date, adjusted, log_return)

head(weekyly_log_returns)
```

